The Iterative Signature Algorithm

### 2009-08-06 isa2/inst/doc/ISA_tutorial.Rnw
isa2 vignette: use Cairo for graphics, if available.
Corrected images in isa2 vignette.

<<Roptions,echo=FALSE,results=hide>>=
## Some options for R
options(width=60)
@

@@

Let us see an example. We load the \Rpackage{isa2} package, and
generate some in 
silico data with the \Rfunction{isa.in.silico} function first:
<<insilico,cache=TRUE>>=
library(isa2)
data <- isa.in.silico(200,200,2,50,50)[[1]]
@
The function call above generates a data matrix similar to the one in
Fig.~\ref{fig:twoclusters}, but a bigger one. The first two arguments define the size of
the data matrix ($200\times 200$), the third the number of modules (two)
and the fourth and fifth give the size of the modules, both are
$50\times 50$ submatrices in this case. By default
\Rfunction{isa.in.silico} creates non-overlapping modules with some
background noise.

@@

\begin{figure}
\centering
\includegraphics[width=0.6\textwidth]{blockplot}
\caption{Another artificial data matrix with two modules and a lot of
  noisy elements.}
\label{fig:block}
\end{figure}

All we have to do now is calling the \Rfunction{isa} function on the
data matrix. 
<<seed,echo=FALSE>>=
## Just to get the same results
set.seed(10)
@ 
<<isa1,cache=TRUE>>=
modules <- isa(data)
names(modules)
@

@@

<<simpleplot,results=hide,include=FALSE>>=
layout(cbind(seq(ncol(modules$rows))))
sapply(seq(ncol(modules$rows)), function(x) {
  par(mar=c(3,5,1,1))
  plot(modules$rows[,x], ylim=c(-1,1), 
       ylab="scores", xlab=NA, 
       cex.lab=2, cex.axis=1.5)
})
@

@@

\begin{figure}
\centering
\includegraphics{tenmods}
\caption{In-silico data with ten non-overlapping modules.}
\label{fig:tenmodules}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=0.6\textwidth]{othermods}
\caption{The input matrix and five modules with relatively low
  robustness scores. As it turns out, each of these modules contains
  two blocks of the input matrix.}
\label{fig:othermods}
\end{figure}

@@

The \Rfunction{drawHeatmap} function of the \Rpackage{biclust} package
can be also used to draw an image plot. This visualizes a single
bicluster on top of the input matrix. It reorders the rows and columns
of the input matrix to move the block of the bicluster to the top
left corner of the input matrix (Fig.~\ref{fig:heatmap}).
<<drawHeatmap.pre,echo=FALSE,results=hide>>=
png("graphics/drawheatmap.png", 400,400)
@ 
<<drawHeatmap>>=
drawHeatmap(five.mods[[1]], Bc, number=1,local=FALSE)
@ 
<<drawHeatmap.post,echo=FALSE,results=hide>>=
dev.off()
@ 
\begin{figure}
\centering
\includegraphics[width=0.5\textwidth]{drawheatmap}
\caption{Heatmap plot of the reordered input matrix. The rows and
  columns corresponding to the first bicluster are moved to the
  top-left corner of the matrix.}
\label{fig:heatmap}
\end{figure}

@@

<<noiseplot.pre,echo=FALSE,results=hide>>=
png("graphics/noise.png", width=800, height=600)
@ 
<<noiseplot>>=
images(lapply(data, "[[", 1), names=as.character(noise), 
       xlab="", ylab="")
@ 
<<noiseplot.post,echo=FALSE,results=hide>>=
dev.off()
@ 
\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{noise}
  \caption{Modular data sets with different levels of noise. The data
  sets are generated by adding normally distributed background noise
  with given variance to the checkerboard matrix.}
  \label{fig:noise}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics{overlapplot}
\caption{The input matrix with two overlapping blocks and the ISA
  modules for this data set. The first three modules were found at
  milder threshold parameters, that is why they are bigger. As you can
  see for Modules \#1 and \#2, at milder thresholds there is a higher
  probability that ISA will pick up some incorrect rows/columns.
}
\label{fig:overlapplot}
\end{figure}

>>>>

<<Roptions,echo=FALSE,results=hide>>=
## Some options for R
options(width=60)
if (require(Cairo)) { png <- CairoPNG }
@

@@

Let us see an example. We load the \Rpackage{isa2} package, and
generate some in 
silico data with the \Rfunction{isa.in.silico} function first:
<<seed,echo=FALSE>>=
## Just to get the same results
set.seed(10)
@ 
<<insilico,cache=TRUE>>=
library(isa2)
data <- isa.in.silico(200,200,2,50,50)[[1]]
@
The function call above generates a data matrix similar to the one in
Fig.~\ref{fig:twoclusters}, but a bigger one. The first two arguments define the size of
the data matrix ($200\times 200$), the third the number of modules (two)
and the fourth and fifth give the size of the modules, both are
$50\times 50$ submatrices in this case. By default
\Rfunction{isa.in.silico} creates non-overlapping modules with some
background noise.

@@

\begin{figure}
\centering
\includegraphics[width=0.6\textwidth]{graphics/blockplot}
\caption{Another artificial data matrix with two modules and a lot of
  noisy elements.}
\label{fig:block}
\end{figure}

All we have to do now is calling the \Rfunction{isa} function on the
data matrix. 
<<isa1,cache=TRUE>>=
modules <- isa(data)
names(modules)
@

@@

<<simpleplot,results=hide,include=FALSE,eval=FALSE>>=
layout(cbind(seq(ncol(modules$rows))))
sapply(seq(ncol(modules$rows)), function(x) {
  par(mar=c(3,5,1,1))
  plot(modules$rows[,x], ylim=c(-1,1), 
       ylab="scores", xlab=NA, 
       cex.lab=2, cex.axis=1.5)
})
@

@@

\begin{figure}
\centering
\includegraphics{graphics/tenmods}
\caption{In-silico data with ten non-overlapping modules.}
\label{fig:tenmodules}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=0.6\textwidth]{graphics/othermods}
\caption{The input matrix and five modules with relatively low
  robustness scores. As it turns out, each of these modules contains
  two blocks of the input matrix.}
\label{fig:othermods}
\end{figure}

@@

The \Rfunction{drawHeatmap} function of the \Rpackage{biclust} package
can be also used to draw an image plot. This visualizes a single
bicluster on top of the input matrix. It reorders the rows and columns
of the input matrix to move the block of the bicluster to the top
left corner of the input matrix (Fig.~\ref{fig:heatmap}).
<<drawHeatmap.pre,echo=FALSE,results=hide>>=
png("graphics/drawheatmap.png", 400,400)
@ 
<<drawHeatmap>>=
drawHeatmap(five.mods[[1]], Bc, number=1,local=FALSE)
@ 
<<drawHeatmap.post,echo=FALSE,results=hide>>=
dev.off()
@ 
\begin{figure}
\centering
\includegraphics[width=0.5\textwidth]{graphics/drawheatmap}
\caption{Heatmap plot of the reordered input matrix. The rows and
  columns corresponding to the first bicluster are moved to the
  top-left corner of the matrix.}
\label{fig:heatmap}
\end{figure}

@@

<<noiseplot.pre,echo=FALSE,results=hide>>=
png("graphics/noise.png", width=800, height=600)
@ 
<<noiseplot>>=
images(lapply(data, "[[", 1), names=as.character(noise), 
       xlab="", ylab="")
@ 
<<noiseplot.post,echo=FALSE,results=hide>>=
dev.off()
@ 
\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{graphics/noise}
  \caption{Modular data sets with different levels of noise. The data
  sets are generated by adding normally distributed background noise
  with given variance to the checkerboard matrix.}
  \label{fig:noise}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics{graphics/overlapplot}
\caption{The input matrix with two overlapping blocks and the ISA
  modules for this data set. The first three modules were found at
  milder threshold parameters, that is why they are bigger. As you can
  see for Modules \#1 and \#2, at milder thresholds there is a higher
  probability that ISA will pick up some incorrect rows/columns.
}
\label{fig:overlapplot}
\end{figure}
